---
title: "QP demo of interior-point solver"
author: "Peter Carbonetto"
date: August 9, 2017
output:
  html_document:
    theme: readable
    include:
      before_body: include/header.html
      after_body: include/footer.html
---

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(
  comment   = "#",
  results   = "hold",
  collapse  = TRUE)
```

## Introduction

This is a small demonstration of the primal-dual interior-point method
for solving a quadratic program with convex objective (*i.e.*,
positive-definite Hessian) and convex, quadratic inequality
constraints:

$$
\begin{array}{ll}
\mbox{minimize}   & x^THx/2 + u^Tx \\
\mbox{subject to} & c_i(x) < b_i, \mbox{for $i = 1, 2,\ldots$}
\end{array}
$$

in which the inequality constraints are quadratic functions,

$$
c_i(x) = x^TP_ix/2 + r_i^Tx.
$$

This particular example comes from:

H. P. Schwefel (1995) Evolution and Optimum Seeking. The minimium
occurs at $(0,1,2,-1)$.

Load the packages and function definitions used in this demo.

```{r load-pkgs, message=FALSE}
library(Matrix)
source("../code/misc.R")
source("../code/ipsolver.R")
```

## Define problem

These variables define the quadratic objective function.

```{r qp-objective}
H <- diag(c(2,2,4,2))
u <- c(-5,-5,-21,7)
```

This data structure specifies the quadratic inequality constraints.

```{r qp-constraints}
constraints <-
  list(c1 = list(P = diag(c(4,2,2,0)),r = c(2,-1,0,-1),b = 5),
       c2 = list(P = diag(c(2,2,2,2)),r = c(1,-1,1,-1),b = 8),
       c3 = list(P = diag(c(2,4,2,4)),r = c(-1,0,0,-1),b = 10))
```

## Solve quadratic program using interior-point method

This function returns the quadratic (L2) norm of x with respect to
matrix A.

```{r qnorm}
qnorm <- function (x, A)
  drop(sqrt(sum(x * (A %*% x))))
```

This function computes $x^TAx/2 + b^Tx + c$.

```{r quadf}
quadf <- function (x, A, b, c = 0)
  qnorm(x,A)^2/2 + sum(b*x) + c
```

This function computes the $m \times n$ Jacobian of the vector-valued
constraint function, and the the $n \times n$ Hessian of the
Lagrangian (minus the Hessian of the objective) for the quadratic
program.

```{r qpjacobian}
qpjacobian <- function (x, z, constraints) {
  n <- length(x)
  m <- length(constraints)
  J <- matrix(0,m,n)
  W <- matrix(0,n,n)
  for (i in 1:m) {
    a     <- constraints[[i]]
    J[i,] <- drop(a$P %*% x + a$r)
    W     <- W + z[i]*a$P
  }
  return(list(J = J,W = W))
}
```

Solve the quadratic program using the primal-dual interior-point
solver.

```{r ip}
options(digits = 4)
out <- ipsolver(x      = rep(0,4),
                obj    = function (x) quadf(x,H,u),
                grad   = function (x) list(g = drop(H %*% x + u),H = H),
                constr = function (x) sapply(constraints,
                                        function (a) with(a,quadf(x,P,r,-b))),
                jac    = function (x, z) qpjacobian(x,z,constraints))
cat("Solution:\n")
print(out$x)
```

Compare the ipsolver output against the analytical solution given above.

## Session information

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
