---
title: "QP demo with linear equality constraints"
author: "Peter Carbonetto"
date: August 21, 2017
output:
  html_document:
    theme: readable
    include:
      before_body: include/header.html
      after_body: include/footer.html
---

```{r knitr, echo=FALSE}
knitr::opts_chunk$set(
  comment   = "#",
  results   = "hold",
  collapse  = TRUE)
```

## Introduction

This is another small demonstration of the primal-dual interior-point
method, similar to the [first demo](qpdemo.html), the main difference
being that this example includes a linear equality constraint. This
example is Problem 14 (QQR-T1-4) from the [Hock & Schittkowski
collection](http://www.klaus-schittkowski.de). The minimium occurs at

$$
((\sqrt{7} - 1)/2, (\sqrt{7} + 1)/4)
$$

Load the packages and function definitions used in this demo.

```{r load-pkgs, message=FALSE}
library(Matrix)
source("../code/misc.R")
source("../code/ipsolver.R")
```

## Define problem

This function returns the quadratic objective function at x.

```{r objective}
obj <- function (x)
  (x[1] - 2)^2 + (x[2] - 1)^2
```

This function returns the gradient and Hessian of the objective
function at x.

```{r gradient-and-Hessian}
grad <- function (x)
  list(g = 2*c(x[1] - 2,x[2] - 1),
       H = diag(c(2,2)))
```
	   
This function returns the value of the quadratic inequality constraint
at x.

```{r constraints}
constr <- function (x)
  x[1]^2/4 + x[2]^2 - 1
```
  
This function computes the 1 x 2 Jacobian of the inequality
constraint function, and the 2 x 2 Hessian of the Lagrangian (minus
the Hessian of the objective).

```{r}
jac <- function (x, z)
  list(J = t(matrix(c(0.5*x[1],2*x[2]))),
       W = z*diag(0.5,2))
```
	   
## Solve quadratic program using IP method

Solve the quadratic program using the primal-dual interior-point
solver. Note that the starting point must be feasible; that is, it
must satisfy the linear equality constraint $Ax = b$ and it must
strictly satisfy the inequality constraint.

```{r ip}
options(digits = 12)
out <- ipsolver(x = c(0.5,0.75),A = t(matrix(c(1,-2))),b = -1,
                obj = obj,grad = grad,constr = constr,jac = jac)
cat("IP solution:\n")
print(out$x)
cat("Compare against the analytic solution:\n")
c((sqrt(7) - 1)/2, (sqrt(7) + 1)/4)
```

Compare the ipsolver output against the analytical solution given above.

## Session information

This is the version of R and the packages that were used to generate
these results.

```{r session-info}
sessionInfo()
```
